### 2015-10-04 mcmc.tex
author affils, small text tweaks

% This file is part of the MCMC project,
%   which is (in turn) part of the Data Analysis Recipes project.
% Copyright 2013, 2014, 2015 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU).

% to-do
% -----
% - mention JAGS and STAN in methods section
% - when we say ``below'' also give a Section Number.
% - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
% - need to mention case where there is an MC integral INSIDE the likelihood function
% - need to mention testing the likelihood function
% - send to friendlies for comments
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
%   - Hill
% - post on arXiv

@@

\noindent
Dan~Foreman-Mackey\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
       New~York~University}\\[1ex]
David~W.~Hogg\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
       New~York~University}\\
\affil{Center~for~Data~Science,
       New~York~University}\\
\affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}

@@

Given this pdf $p(\pars)$, we can define \emph{expectation values} $E[\pars]$
for $\pars$ or for any quantity that can be expressed as a function $q(\pars)$
of $\pars$:
\begin{eqnarray}
E[\pars] &\equiv& \int \pars\,p(\pars)\,\dd \pars
\\
E[q] &\equiv& \int q(\pars)\,p(\pars)\,\dd \pars
\quad ,
\end{eqnarray}
where again the integrals are implicitly definite integrals over the
entire domain
of $\pars$ (all the parts of $\pars$ space in which $p(\pars)$ is finite) and the
integrals are multi-dimensional if $\pars$ is multi-dimensional.
These expectation values are the mean values of $\pars$ and $q(\pars)$ under
the pdf.  A good sampling---really the definition of a good
sampling---makes the sampling approximation to these integrals
accurate.
With a good sampling $\setof{\pars_k}_{k=1}^K$ the integrals get replaced
with sums over samples $\pars_k$:
\begin{eqnarray}
E[\pars] &\approx& \frac{1}{K}\,\sum_{k=1}^K \pars_k
\\
E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
\quad .
\end{eqnarray}
That is, a sampling is good when any expectation value of interest is
accurately computed via the sampling approximation.
The word ``accurately'' here translates into some kinds of theorems
about limiting behavior; the general idea is that the sampling
approximation becomes exact as $K$ goes to infinity.
The size $K$ of the sampling you need \emph{in practice} will depend
on the expectations you want to compute, and the accuracies you need.

@@

\section{What is Metropolis--Hastings MCMC?}

@@

The autocorrelation function is defined by
\begin{eqnarray}
C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
\label{eq:autocorrelationfn}\\
\mu &=& E[\pars]
\quad ;
\end{eqnarray}
that is, it is the covariance of points separated by $\Delta$, taken
relative to the variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the MCMC algorithm.
By construction
\begin{eqnarray}
C(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau = \sum_{\Delta = 0}^{\infty} C(\Delta)
\label{eq:autocorrelationtime}\quad .
\end{eqnarray}
Heuristically, the autocorrelation time is the number of samples you
need to take before you get an independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.\note{HOGG Explain more fully.  Maybe also cite Fengji.}
It is slightly off-topic here, but related to this, if you have a long
chain of length $K$ and with autocorrelation time $\tau$ and you thin
the chain down to every $\tau$th sample (instead of using all $K$
samples), you will get just as good sampling-based estimates as if you
had used all $K$ samples.

A sampler with a better autocorrelation time is just better; you have
to do fewer $f(\pars)$ calls per independent sample, and you have to
run less time to get accurate sampling-based integral estimates.
A sampler that takes an independent sample every time would have an
autocorrelation time of unity, which is the best possible value; this
optimal sampling is only possible for problems where the sampling is
analytic (for example if the posterior is perfectly Gaussian with
known mean and variance).
In general, the best sampler will be different for different problems,
and we will (below) tune the samplers we have to do the best on the
problems we have; this tuning will also be problem-specific.
There are many heuristic bases on which different samplers might be
compared or tuned, but fundamentally it is lower autocorrelation time
that separates good samplers from bad ones and is the ultimate basis
on which we compare performance.\note{HOGG: Bash Dunkley and anyone
  else who has used something else to compare.}

@@

%% DWH:  Note here that this requires writing the f(\pars) code to capitalize on the axis-aligned moves!!

@@

% Hogg say:  I have no idea what the comment here (below) means, even though I wrote it!
% ---
% Give a sense of the difficulties you get into if you fail here; like
% discussion with Sven Kreiss on 2013-02-14...
% ---

>>>>

% This file is part of the MCMC project,
%   which is (in turn) part of the Data Analysis Recipes project.
% Copyright 2013, 2014, 2015 David W. Hogg (NYU) and Dan Foreman-Mackey (NYU, UW).

% to-do
% -----
% - deal with all mentions of ``DFM'' or ``HOGG''
% - mention Gelman--Rubin statistic
% - mention JAGS and STAN in methods section
% - when we say ``below'' also give a Section Number.
% - where to put comments about how to structure code ln_likelihood(pars, data, I) and ln_prior(pars, I)
% - need to mention case where there is an MC integral INSIDE the likelihood function
% - need to mention testing the likelihood function
% - send to friendlies for comments
%   - Gelman
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - post on arXiv

@@

\noindent
Dan~Foreman-Mackey\\
\affil{Department~of~Astronomy, University~of~Washington}\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
       New~York~University}\\[1ex]
David~W.~Hogg\\
\affil{Simons~Center~for~Data~Analysis,
       Simons~Foundation}\\
\affil{Center~for~Cosmology~and~Particle~Physics, Department~of~Physics,
       New~York~University}\\
\affil{Center~for~Data~Science,
       New~York~University}\\
\affil{Max-Planck-Institut~f\"ur~Astronomie, Heidelberg}

@@

Given this pdf $p(\pars)$, we can define \emph{expectation values} $E[\pars]$
for $\pars$ or for any quantity that can be expressed as a function $q(\pars)$
of $\pars$:
\begin{eqnarray}
E[\pars] &\equiv& \int \pars\,p(\pars)\,\dd \pars
\\
E[q] &\equiv& \int q(\pars)\,p(\pars)\,\dd \pars
\quad ,
\end{eqnarray}
where again the integrals are implicitly definite integrals over the
entire domain
of $\pars$ (all the parts of $\pars$ space in which $p(\pars)$ is finite) and the
integrals are multi-dimensional if $\pars$ is multi-dimensional.
These expectation values are the mean values of $\pars$ and $q(\pars)$ under
the pdf.  A good sampling---and really this is the definition of a good
sampling---makes the sampling approximation to these integrals
accurate.
With a good sampling $\setof{\pars_k}_{k=1}^K$ the integrals get replaced
with sums over samples $\pars_k$:
\begin{eqnarray}
E[\pars] &\approx& \frac{1}{K}\,\sum_{k=1}^K \pars_k
\\
E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(\pars_k)
\quad .
\end{eqnarray}
That is, a sampling is good when any expectation value of interest is
accurately computed via the sampling approximation.
The word ``accurately'' here translates into some kinds of theorems
about limiting behavior; the general idea is that the sampling
approximation becomes exact as $K$ goes to infinity.
The size $K$ of the sampling you need \emph{in practice} will depend
on the expectations you want to compute, and the accuracies you need.

@@

\section{Metropolis--Hastings MCMC}

@@

The autocorrelation function is defined by
\begin{eqnarray}
C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
\label{eq:autocorrelationfn}\\
\mu &=& E[\pars]
\quad ;
\end{eqnarray}
that is, it is the covariance of points separated by $\Delta$, taken
relative to the variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the MCMC algorithm.
By construction
\begin{eqnarray}
C(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau = \sum_{\Delta = 0}^{\infty} C(\Delta)
\label{eq:autocorrelationtime}\quad .
\end{eqnarray}
%% DFM: IS THIS DEFINITION CORRECT OR IS THERE A FACTOR OF 2 OR SOMETHING?
Heuristically, the autocorrelation time is the number of samples you
need to take before you get an independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.\note{HOGG Explain more fully.  Maybe also cite Fengji.}
It is slightly off-topic here, but related to this, if you have a long
chain of length $K$ and with autocorrelation time $\tau$ and you thin
the chain down to every $\tau$th sample (instead of using all $K$
samples), you will get just as good sampling-based estimates as if you
had used all $K$ samples.

A sampler with a smaller autocorrelation time is just better; you have
to do fewer $f(\pars)$ calls per independent sample, and you have to
run less time to get accurate sampling-based integral estimates.
A sampler that takes an independent sample every time would have an
autocorrelation time of unity, which is the best possible value; this
optimal sampling is only possible for problems where the sampling is
analytic (for example if the posterior is perfectly Gaussian with
known mean and variance).
In general, the best sampler will be different for different problems,
and we will (below) tune the samplers we have to do the best on the
problems we have; this tuning will also be problem-specific.
There are many heuristic bases on which different samplers might be
compared or tuned, but fundamentally it is lower autocorrelation time
that separates good samplers from bad ones and is the ultimate basis
on which we compare performance.\note{HOGG: Bash Dunkley and anyone
  else who has used something else to compare.}

@@

%% HOGG:  Note here that this requires writing the f(\pars) code to capitalize on the axis-aligned moves!!

@@

% HOGG say:  I have no idea what the comment here (below) means, even though I wrote it!
% ---
% Give a sense of the difficulties you get into if you fail here; like
% discussion with Sven Kreiss on 2013-02-14...
% ---
